Retrofitting Effect Handlers onto OCaml

### 2020-12-10 retro-concurrency.tex
Make all the fixes pointed out by Francois

The interaction of non-local control flow with systems programming is quite
subtle~\cite{TFP17, Leijen18}. Consider the following function that uses
blocking I/O functions from the OCaml standard library to copy data from the
input channel |ic| to the output channel |oc|:
\begin{lstlisting}
let rec copy ic oc =
  try output_string oc ((input_line ic) ^ "\n");
      copy ic oc
  with End_of_file -> close_in ic; close_out oc
	   | e -> close_in ic; close_out oc; raise e
\end{lstlisting}
The function |input_line| raises an |End_of_file| exception on reaching the end
of input, which is handled by the exception handler which closes the channels.
The |close_*| functions do nothing if the channel is already closed. The code
is written in a defensive style to handle other exceptional cases such as the
channels being closed externally. Both |input_line| and |output_string| raise
|Sys_error| exception if the channel is closed. In this case, the catch-all
exception handler closes the channels and reraises the exception to communicate
the exceptional behaviour to the caller.

One of our goals (\S\ref{sec:req}) is to make this code transparently
asynchronous. We can define effects for performing the I/O operations and wrap
them up in functions with the same signature as the one from the standard
library:
\begin{lstlisting}
effect In_line : in_channel -> string
effect Out_str : out_channel -> string -> unit
let input_line ic = perform (In_line ic)
let output_string oc = perform (Out_str oc)
\end{lstlisting}

@@

In the operational semantics, the continuations may be resumed more than once.
Captured continuations are copies of the original fibers and resuming the
continuation copies the fibers and leaves the continuation as it is. In our use
cases, continuations will be resumed at-most once, and copying fibers is
unnecessary and inefficient. Instead, Multicore OCaml optimises fibers for
one-shot continuations. Fibers are allocated on the heap using |malloc| and are
|free|d when no longer necessary. Similar to Farvardin et
al.~\cite{Farvardin20}, we use a \emph{stack cache} of recently freed stacks in
order to speed up allocation.

@@

When the effect |E| is performed, we allocate a continuation object |ke| in the
heap that points to the current fiber |f|, set fiber |f|'s parent pointer to
|NULL|, and evaluate the continuation closure |clos_heffect| on the parent
fiber |e| with the effect |E| and the continuation |ke| as arguments. Since the
first handler does not handle effect |E|, the effect is \emph{reperformed}
(\S\ref{sec:static_semantics}) by appending the fiber |e| to the tail of
continuation |ke|, set fiber |e|'s parent pointer to |NULL|, and evaluate the
current continuation closure on the parent fiber |main| with |E| and |ke| as
arguments, which handles |E| (position |p2|). Thus, \emph{continuations are
captured without copying frames}. Since every handler closure is evaluated
until a matching one is found, the time taken to handle an effect is linear in
the number of handlers. We observed that the handler stack is shallow in
real programs.

When the continuation is resumed, we overwrite the value of |ke| to |NULL| to
enforce at-most once semantics. Resuming a continuation involves traversing the
linked-list of fibers and making the last fiber point to the current fiber.
Just as in the operational semantics, the implementation invokes the
appropriate closure to either |continue| or |discontinue| the continuation
(position |p3|). We perform standard tail-call optimisation so that resumptions
at tail positions do not build up stack.

@@

There are several strategies for implementing effect handlers. Eff~\cite{Eff},
Helium~\cite{Biernacki20}, Frank~\cite{Frank} and the Links server
backend~\cite{Hillerstrom20} use an interpreter similar to our operational
semantics to implement effect handlers. Effekt~\cite{Effekt}, Links JavaScript
backend~\cite{Hillerstrom20} and Koka~\cite{Leijen17} use type-directed
selective CPS translation. These language are equipped with an effect system,
which allows compiling pure code in direct-style and effectful code in CPS.
Leijen~\cite{Leijen14} implements effect handlers as a library in C using stack
copying. C allows pointers into the stack, so care is taken to ensure that
when continuations are resumed, the constituent frames are restored to the same
memory addresses as at the time of capture. Kiselyov et al.~\cite{Kiselyov18}
use an implementation of multi-prompt delimited continuations as an OCaml
library~\cite{Kiselyov12} to embed the Eff language in OCaml. Indeed, Forester et
al.~\cite{Forster19} showed that in an untyped setting, effect handlers,
monadic reflection and delimited control can macro-express each other.

>>>>

The interaction of non-local control flow with systems programming is quite
subtle~\cite{TFP17, Leijen18}. Consider the following function that uses
blocking I/O functions from the OCaml standard library to copy data from the
input channel |ic| to the output channel |oc|:
\begin{lstlisting}
let copy ic oc =
  let rec loop () =
	  output_string oc ((input_line ic) ^ "\n");
		loop ()
	in
	try loop () with
  | End_of_file -> close_in ic; close_out oc
	| e -> close_in ic; close_out oc; raise e
\end{lstlisting}
The function |input_line| raises an |End_of_file| exception on reaching the end
of input, which is handled by the exception handler which closes the channels.
The |close_*| functions do nothing if the channel is already closed. The code
is written in a defensive style to handle other exceptional cases such as the
channels being closed externally. Both |input_line| and |output_string| raise
|Sys_error| exception if the channel is closed. In this case, the catch-all
exception handler closes the channels and reraises the exception to communicate
the exceptional behaviour to the caller.

One of our goals (\S\ref{sec:req}) is to make this code transparently
asynchronous. We can define effects for performing the I/O operations and wrap
them up in functions with the same signature as the one from the standard
library:
\begin{lstlisting}
effect In_line : in_channel -> string
effect Out_str : out_channel * string -> unit
let input_line ic = perform (In_line ic)
let output_string oc s = perform (Out_str (oc, s))
\end{lstlisting}

@@

In the operational semantics, the continuations may be resumed more than once.
Captured continuations are copies of the original fibers and resuming the
continuation copies the fibers and leaves the continuation as it is. In our use
cases, continuations will be resumed at-most once, and copying fibers is
unnecessary and inefficient. Instead, Multicore OCaml optimises fibers for
one-shot continuations. Fibers are allocated on the C heap using |malloc| and
are |free|d when the handled computation returns with a value or an exception.
Similar to Farvardin et al.~\cite{Farvardin20}, we use a \emph{stack cache} of
recently freed stacks in order to speed up allocation.

@@

When the effect |E| is performed, we allocate a continuation object |ke| in the
OCaml heap that points to the current fiber |f|, set fiber |f|'s parent pointer
to |NULL|, and evaluate the continuation closure |clos_heffect| on the parent
fiber |e| with the effect |E| and the continuation |ke| as arguments. Since the
first handler does not handle effect |E|, the effect is \emph{reperformed}
(\S\ref{sec:static_semantics}) by appending the fiber |e| to the tail of
continuation |ke|, set fiber |e|'s parent pointer to |NULL|, and evaluate the
current continuation closure on the parent fiber |main| with |E| and |ke| as
arguments, which handles |E| (position |p2|). Thus, \emph{continuations are
captured without copying frames}. Since every handler closure is evaluated
until a matching one is found, the time taken to handle an effect is linear in
the number of handlers. We observed that the handler stack is shallow in real
programs.

When the continuation is resumed, we overwrite the value of |ke| to |NULL| to
enforce at-most once semantics. Resuming a continuation involves traversing the
linked-list of fibers and making the last fiber point to the current fiber.
Just as in the operational semantics, the implementation invokes the
appropriate closure to either |continue| or |discontinue| the continuation
(position |p3|). We perform tail-call optimisation so that resumptions at tail
positions do not build up stack.

@@

There are several strategies for implementing effect handlers. Eff~\cite{Eff},
Helium~\cite{Biernacki20}, Frank~\cite{Frank} and the Links server
backend~\cite{Hillerstrom20} use an interpreter similar to our operational
semantics to implement effect handlers. Effekt~\cite{Effekt}, Links JavaScript
backend~\cite{Hillerstrom20} and Koka~\cite{Leijen17} use type-directed
selective CPS translation. These language are equipped with an effect system,
which allows compiling pure code in direct-style and effectful code in CPS.
Leijen~\cite{Leijen14} implements effect handlers as a library in C using stack
copying. C allows pointers into the stack, so care is taken to ensure that
when continuations are resumed, the constituent frames are restored to the same
memory addresses as at the time of capture. Kiselyov et al.~\cite{Kiselyov18}
use an implementation of multi-prompt delimited continuations as an OCaml
library~\cite{Kiselyov12} to embed the Eff language in OCaml. Indeed, Forster et
al.~\cite{Forster19} showed that in an untyped setting, effect handlers,
monadic reflection and delimited control can macro-express each other.
